---
title: covid-19 situation in Phillipines
author: Brenwin
date: '2020-09-16'
slug: covid-19-situation-in-phillipines
categories:
  - covid19
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-09-16T21:43:24+10:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
bibliography: [reference.bib]
link-citations: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plotly)
library(crosstalk)
library(lubridate)
library(colorspace)
library(scales)
library(patchwork)
```

# About Data
Data set was retrieved from Department of Health in Phillipines (@DOH). It is a very comprehensive data set which includes case information, hospital reports and quarantine facilities report among others.  

# Overview of Coronavirus cases in the world's most stringent lockdown country

```{r readin-data}
#--- save data sets to binary formats 
# case_info <- read_csv("data/case-info.csv")
# save(case_info, file = "data/case_info.rda")

# daily_report <- read_csv("data/daily-report.csv")
# save(daily_report, file = "data/daily_report.rda")

# load in R
load("data/case_info.rda") 
load ("data/daily_report.rda")

testing_agg <- read_csv("data/testing-aggregates.csv")

base_data <- read_csv("data/base-data.csv")
```

```{r case-info-wrangling}
# select relevant variables
case_info_clean <- case_info %>% 
  select(AgeGroup, Sex, DateRepConf, DateDied, DateRecover, RemovalType, Admitted, RegionRes, ProvRes, HealthStatus, Quarantined)

# daily cases
case_info_clean <- case_info_clean %>%
  group_by(DateRepConf) %>% 
  mutate(cases = n()) 

# daily recovered  
case_info_clean <- case_info_clean %>% 
  ungroup() %>% 
  group_by(DateRecover) %>% 
  mutate(recovered = n()) %>% 
  # remove NA 
  mutate(recovered = ifelse(recovered %in% 228796,
                            yes = NA,
                            no = recovered))


# daily deaths
case_info_clean <- case_info_clean %>% 
  ungroup() %>% 
  group_by(DateDied) %>% 
  mutate(deaths = n()) %>% 
  # remove NA 
  mutate(deaths = ifelse(deaths %in% 271554,
                         yes = NA,
                         no = deaths)) %>% 
  ungroup()

```

```{r wrangle-plot1}
# fill in missing dates
case_info_clean1 <- case_info_clean %>% 
  tidyr::complete(DateRepConf = seq.Date(from = min(DateRepConf),
                           to = max(DateRepConf),
                           by = "day"))

case_info_clean1 <- case_info_clean1 %>% 
  distinct(DateRepConf, .keep_all = TRUE) %>% 
  pivot_longer(cols = cases:deaths,
               names_to = "type",
               values_to = "count") %>% 
  mutate(count = ifelse(test = is.na(count),
                        yes = 0,
                        no = count))

case_info_clean1 <- case_info_clean1 %>%
  group_by(RegionRes, type) %>%
  mutate(cum_count = cumsum(count))

# rename variables
case_info_clean1 <- case_info_clean1 %>% 
  rename(Date = DateRepConf,
         Region = RegionRes)
```

```{r overall-plot}
color_ph <- c("#0a0af5", # blue
           "#f50a0a", # red
           "#f5f50a") # yellow

theme_ph <- theme(
  #background & panels
  plot.background = element_rect(fill = "#87adf1"),
  panel.background = element_rect(fill = "#87adf1"),
  # customize facet strip
  strip.background = element_rect(fill = "#87adf1"),
  strip.text.x = element_text(size = 14, colour = "white"),
  
  # customize plot title
  plot.title = element_text(size = 22, face = "bold", colour = "white"),
  
  # axes title and tick labels
  axis.title.y = element_text(colour = "white", size = 13),
  axis.title.x = element_text(colour = "white", size = 13),
  axis.text = element_text(colour = "white", size = 11)
)


# cases
p1_cases <- case_info_clean1 %>% 
  filter(type == "cases") %>% 
  ggplot() +
  geom_line(aes(x = Date,
                y = count),
            colour = "#0a0af5") +
  labs(tags = "cases") +
  scale_x_date(date_breaks = "1 month",
               date_minor_breaks = "2 week",
               date_labels = format("%b")) +
  annotate(geom = "label",
           x = as.Date("2020-01-30", "%Y-%m-%d"),
           y = 800,
           label = "first case") +
  annotate(geom = "label",
           x = as.Date("2020-03-17", "%Y-%m-%d"),
           y = 800,
           label = "lockdown begins") +
  annotate(geom = "label",
           x = as.Date("2020-06-01", "%Y-%m-%d"),
           y = 1500,
           label = "restrictions eased") +
  annotate(geom = "label",
           x = as.Date("2020-08-14", "%Y-%m-%d"),
           y = 7500,
           label = "renew strict lockdown") +
  theme_ph +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()) 

p1_cases_ply <- ggplotly(p1_cases)

# deaths
p2_deaths <- case_info_clean1 %>% 
  filter(type == "deaths") %>% 
  ggplot() +
  geom_line(aes(x = Date,
                y = count),
            colour = "#f50a0a") +
  labs(tags = "deaths") +
  scale_x_date(date_breaks = "1 month",
               date_minor_breaks = "2 week",
               date_labels = format("%b")) +
  theme_ph +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

p2_deaths_ply <- ggplotly(p2_deaths)

# recovered
p3_recovered <- case_info_clean1 %>% 
  filter(type == "recovered") %>% 
  ggplot() +
  geom_line(aes(x = Date,
                y = count),
            colour = "#f5f50a") +
  labs(tags = "recovered") +
  scale_x_date(date_breaks = "1 month",
               date_minor_breaks = "2 week",
               date_labels = format("%b")) +
  theme_ph +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())

p3_recovered_ply <- ggplotly(p3_recovered)

p1_cases / (p2_deaths + p3_recovered) + 
  patchwork::plot_layout(widths = c(3,3,3))
```

The first case of coronavirus in Philippines was recorded on 30th January 2020. Philippines was the first country outside of China to record a confirmed death from the virus (@nytimes). They did not see another case for a month before it met its first local transmission on March 7 as reported on @phcnn. 

Since then, the virus has been spread to every region around Phillipines. (you might want to play around with interactive region plot below!). With a surge in cases, in about 10 days, what has been touted the "strictest coronavirus lockdown in the world" was enforced. With the President ordered to "shoot dead" whoever who defied the lockdown; emulating Philippines tactics in the drug war Philippines was still recovering from. (@cbsnews) Involving house to house visits, a mandated 8pm-5am curfew extending to over 60million people, deploying army and police to man checkpoints.(@lowly-inst). 

```{r region-plot-wrangle}
theme_ph <- theme_ph +
  theme(legend.background = element_rect(fill = "#87adf1"))
case_info_ct <- highlight_key(case_info_clean1)

# cases plot
p1 <-  ggplot(case_info_ct) +
  geom_line(aes(x = Date,
               y = cum_count,
               group = Region,
               colour = Region)) +
  scale_x_date(date_breaks = "1 month",
               date_minor_breaks = "2 week",
               date_labels = format("%b")) +
  labs(x = "Date",
       y = "Cumulative Count",
       title = "Region Cumulative") +
  facet_wrap(~type) +
  guides(colour = FALSE) +
  colorspace::scale_color_discrete_qualitative(palette = "Dark 2") +
  theme_ph +
  theme(plot.title.position = "plot",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 90)) 
  
```

```{r region-plot}
ui_controls <- bscols(
  filter_select("id", "Select a Region", case_info_ct, ~Region),
  ggplotly(p1) %>% 
    layout(xaxis = list(automargin = TRUE),
           yaxis = list(automargin = TRUE)) %>% 
    config(displayModeBar = F) %>% 
                    layout(xaxis = list(fixedrange = TRUE),
                           yaxis = list(fixedrange = TRUE)) %>% 
    hide_legend(),
  widths = c(10, 10)
)

bscols(ui_controls)
```
  
  
Despite the strict lockdown, we see an exponential rise in cases. In many articles, the socio-economic status has been associated with the prevalence of the covid-19 cases. (@lowly-inst) It even prompted UN (& partners) towards a $5.4 million Covid-19 response plan to aid poor & marginilized community- one of the largest humanitarian aid since Typhoon Haiyan/Yolanda. (@un-plan)

Between March 17 & July 25 a baffling 76,000 arrests were made. The pandemic had devastating effects espcially on the poor- exposing the gaping inequalities in access to food, water and shelter. (@cbsnews) With the data, it was found that most of these communities were actually in the city regions such as ROF and Metro Manila(NCR)   

# Rising cases of coronavirus. Is easing restrictions going to be the best way forward?

Philippines is currently contemplating easing its restrictions despite having the highest number of cases in South-East Asia. (@st). Initially, there was a call from senior doctors to imposing a stricter lockdown in Manila- a much needed "timeout" for the overwhelming hospitals. (@guardian) Data below examines the demographics and current state of capacity. 

```{r dem-plot}
# replace NA values in AgeGroup; to unknown
dem_info <- case_info_clean %>% 
  mutate(AgeGroup = replace_na(AgeGroup, "Unknown"))

dem_info$AgeGroup <- factor(dem_info$AgeGroup,
                                   levels = c("0 to 4", 
                                              "5 to 9",
                                              "10 to 14",
                                              "15 to 19",
                                              "20 to 24",
                                              "25 to 29",
                                              "30 to 34",
                                              "35 to 39",
                                              "40 to 44",
                                              "45 to 49",
                                              "50 to 54",
                                              "55 to 59",
                                              "60 to 64",
                                              "65 to 69",
                                              "70 to 74",
                                              "75 to 79",
                                              "80+",
                                              "Unknown"))

# cases 
dem_cases <- dem_info %>% 
  ungroup() %>% 
  group_by(AgeGroup, Sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(type = "cases")

# deaths
dem_deaths <- dem_info %>% 
  group_by(AgeGroup, Sex) %>% 
  filter(RemovalType == "DIED") %>% 
  count() %>% 
  mutate(type = "deaths")

# recovered
dem_recovered <- dem_info %>% 
  group_by(AgeGroup, Sex) %>% 
  filter(RemovalType == "RECOVERED") %>% 
  count() %>% 
  mutate(type = "recovered")
  

# combine rows
dem <- rbind(dem_cases, dem_deaths, dem_recovered) %>% 
  ungroup() %>% 
  group_by(type) %>% 
  mutate(total = sum(n)) %>% 
  ungroup() %>% 
  rowwise() %>% 
  # Age Group & sex % in each type (cases// deaths// recovered)
  mutate(pct = n/total)
  

# specify colours
dem_plot_col <- c("#bf5d8c", "#0037b4")
  
dem_plot <- dem %>% 
  ggplot() +
  geom_col(aes(x = AgeGroup,
               y = pct,
               fill = Sex,
               label = n,
               text = paste('</br> Age Group:', AgeGroup,
                            '</br> Percentage:', scales::percent(dem$pct,
                                                                 accuracy = 0.01),
                            '</br> Absolute Number:', n),
               position = "dodge",
               stat = "identity")) +
  scale_fill_manual(values = dem_plot_col) +
  coord_flip() + 
  facet_wrap(~type, 
             ncol = 3,
             scales = "free_x") +
  labs(x = "AgeGroup",
       y = "percent") +
  theme_ph 
  

ggplotly(dem_plot, 
         tooltip = c("text"))
```

The interactive plot above shows demographics of cases as a whole. While most cases were in the middle-aged, we see majority of the deaths were coming from 60+ population. No stark difference is seen in males and females.

```{r}
test_plot_col <- c("#0037b4", "#bf5d8c")

# testing per day
testing_clean <- testing_agg %>% 
  group_by(report_date) %>% 
  summarize(daily_tests = sum(daily_output_samples_tested, na.rm = TRUE),
         positive = sum(daily_output_positive_individuals, na.rm = TRUE)) %>% 
  pivot_longer(cols = daily_tests:positive,
               names_to = "test_results",
               values_to = "values") 

test_bar <- testing_clean %>% 
  ggplot(aes(x = report_date,
             y = values,
             fill = test_results)) +
  scale_fill_manual(values = test_plot_col) +
  geom_col() +
  labs(x = "Report Date",
       y = "Individuals Sampled",
       tag = "Daily Samples Tested") +
  theme_ph

test_rate_line <- testing_clean %>% 
  pivot_wider(names_from = test_results,
              values_from = values) %>% 
  mutate(daily_postv_rate = positive/daily_tests) %>% 
  ggplot() +
  geom_line(aes(x = report_date,
                y = daily_postv_rate),
            colour = "#bf5d8c") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_date(date_breaks = "1 month",
               date_minor_breaks = "2 week",
               date_labels = format("%b")) +
  labs(x = "Date",
       y = "daily positive rate",
       tag = "Daily Positive Rate") +
  theme_ph

p1_test <- ggplotly(test_bar) %>% 
  layout(title = "Daily Samples Tested")
  

p2_test <- ggplotly(test_rate_line)

plotly::subplot(test_bar, test_rate_line,
                shareX = TRUE,
                margin = 0.10)
```

A rising amount of testing were conducted from 118 testing facilities now in place. The positivity in testing results shows an undulating trend but within a 5% to 15% trend and seems to be on its way down. An easing of restrictions might or might bring them back to its original state. 

```{r hospital, include = FALSE}
# https://www.doh.gov.ph/2019-nCov

# availability of beds
hos1 <- tribble(~type, ~vacancy, ~number,
        "ICU Beds", "vacant", 1025,
        "ICU Beds", "occupied", 881,
        "Isolation Beds", "vacant", 7650,
        "Isolation Beds", "occupied", 6186,
        "Ward Beds", "occupied", 2640,
        "Ward Beds", "vacant", 3140) %>% 
  ggplot() +
  geom_col(aes(x = type,
               y = number,
               fill = vacancy)) +
  scale_fill_manual(values = c("#bf5d8c", "#0037b4")) +
  coord_flip() +
  ggtitle("Hospital Beds Occupancy") +
  theme_ph +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank())

hos1_ply <- ggplotly(hos1)

hos2 <- tribble(~type, ~availability, ~number,
        "Mechanical ventilators", "occupied", 586,
        "Mechanical ventilators", "vacant", 1616) %>% 
  ggplot() +
  geom_col(aes(x = type,
               y = number,
               fill = availability)) +
  scale_fill_manual(values = c("#bf5d8c", "#0037b4")) +
  coord_flip() +
  ggtitle("Availability of equipment") +
  theme_ph +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank()) 

hos2_ply <- ggplotly(hos2)

hos3 <- tribble(~type, ~number,
        "Safe(<30% occupied)", 1025,
        "Warning (30% to 70%)", 294,
        "Danger(>70%)", 239) %>% 
  ggplot() +
  geom_col(aes(x = fct_reorder(type,number),
               y = number,
               fill = number)) +
  xlab("type") +
  colorspace::scale_fill_continuous_qualitative(palette = "Cold") +
  coord_flip() +
  ggtitle("Hospital Occupancy Rate level") +
  theme_ph + 
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor.x = element_blank()) +
  guides(fill = FALSE) 

hos3_ply <- ggplotly(hos3)

(hos1 + hos3) / hos2 +
  plot_layout()
```

```{r}
hos3_ply
```

```{r}
hos1_ply
```



```{r}
hos2_ply
```

In terms of hospital capacity and availability of equipment exclusively for Covid-19 patients (from @DOH), it is a very contentious decision to make as the government need to strike between reviving the economy and safety.  

However, considering the number of new cases are slowly decreasing it may not be a bad idea to start gradually easing the restrictions. 

# Acknowledgements
All of these are done using R software  
```{r, include = FALSE}
citation("rmarkdown") %>%  toBibtex(.)
citation("tidyverse")
citation("plotly")
citation("crosstalk")
citation("lubridate")
citation("colorspace")
citation("scales")
citation("patchwork")
citation("blogdown") %>% toBibtex(.)
```

Including: these R packages are used:  
- `Rmarkdown`: @Rmd  
- `tidyverse`: @Rtidyverse    
- `plotly`: @Rplotly  
- `crosstalk`: @Rcrosstalk  
- `lubridate`: @Rlubridate  
- `colorspace`: @Rcolorspace  
- `scales`: @Rscales  
- `patchwork`: @Rpatchwork
- `blogdown`: @Rblogdown


# References




